---
title: "Viz Makeover: Survey of Public Transit Agencies"
categories:
  - viz
description: |
  Making over a viz from APTA
date: 09-19-2020
output:
  distill::distill_article:
    self_contained: false
---
Context and audience mean so much for data visualization. On Twitter, I sometimes see charts that have great information but aren't optimized for the Twitter experience. Some common missed opportunities:

* They're too big, so the Twitter crop hides the important parts

* The title doesn't give the big idea, so it's easy to scroll on by

* The viz doesn't take advantage of branding to associate the analysis with the organization

Sometimes I'm inspired to make charts over to be more Twitter-friendly, mostly for my own fun. Here's one I did using a survey of American public transit agencies by the American Public Transportation Association.

## Setup

I'm only using three packages today - `tidyverse` as the anchor, `showtext` for custom fonts, and `ggtext` for colorizing the category labels.

The [original viz](https://www.apta.com/wp-content/uploads/APTA-Brief-Agency-Survey-Sept-2020.pdf) had the data points in the image, so I made a quick `tribble` to capture the data.

```{r}
library(tidyverse)
library(showtext)
library(ggtext)

#Fonts
font_add_google(name = "Roboto", family = "roboto")
font_add_google(name = "Rubik", family = "rubik")
showtext_auto()

# APTA Data

apta_data <- tibble::tribble(
                                    ~category, ~value,
                           "Eliminate Routes",   0.38,
                     "Reduce Days of Service",   0.20,
                                "Cut Service",   0.61,
                             "Furlough Staff",   0.23,
                              "Lay Off Staff",   0.31,
                             "Increase Fares",   0.17,
  "Delay, Defer, or Cancel Vehicle Purchases",   0.33,
   "Delay, Defer, or Cancel Capital Projects",   0.45
  )

```

## Organize data

I love using `ggtext` to make the category labels pop, but it takes a tiny bit of set up. 

* First, I name all the colors, which I copied from the APTA website to match the branding. 

* Then, I `mutate()` to make a `highlight` column to specify the color. 

* A second `mutate()` plus some basic HTML makes a new `label` column that will be called in the chart. 

* I also `fct_reorder()` the data based on percentage value.

```{r}

# colors
bg_gray <- "#F5F5F5"
axis_line_gray <- "#B8B8B8"
other_gray <- "#474747"

highlight_1 <- "#650533" #apta maroon
highlight_2 <- "#094279" #apta blue

#add in new columns re: color and labels

plot_data <- apta_data %>%
  mutate(
    highlight = case_when(
      category == "Cut Service" ~ highlight_1,
      str_detect(category, "Capital") ~ highlight_2,
      TRUE ~ 	other_gray
    ),
    category_label = 
      glue::glue("<span style ='color:{highlight}'>{category}</span>"),
    category_label = fct_reorder(category_label, value)
  )

```

## Plotting

Now it's time to make a chart! 

* I kept the column chart style and kept the `value` on the x-axis. 

* The y-axis became the `category_label` and requires `element_markdown()` in the theme to render the HTML.

* Then, for the fill, I used the `highlight` column along with `scale_fill_identity()`. The bars also print from largest to smallest value because of the factor created in the step above.

* A big title and subtitle explain the chart. It took some trial and error to get the right size font plus using `<br>` for some manual spacings. These also require `element_markdown()` in the theme.

* I use some custom Google fonts to spice up the default san serif.

```{r}

# the chart!
new_plot <- plot_data %>%
  ggplot(aes(x = value, y = category_label, fill = highlight)) +
  geom_col() +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_fill_identity() +
  labs(title = glue::glue("Without emergency funding from Congress, 
  your transit agency<br>might <span style='color:{highlight_1}'>cut 
  service</span> or <span style='color:{highlight_2}'>delay 
  infrastructure projects</span>"),
       subtitle = "Percent of transit agencies that said they were 
                  considering the following actions",
       caption = "n = 128<br>Source: APTA.com, Sep 2020") +
  theme_classic(base_size = 12) +
  theme(
    #align title left
    plot.title.position = "plot",
    #colors and fonts
    text = element_text(family = "rubik"),
    plot.background = element_rect(fill = bg_gray),
    panel.background = element_rect(fill = bg_gray,
                                    colour = bg_gray),
    axis.line = element_line(color = axis_line_gray),
    plot.title = element_markdown(family = "roboto", 
                                  face = "bold"),
    plot.subtitle = element_markdown(family = "roboto"),
    plot.caption = element_markdown(), 
    axis.text.y = element_markdown(),
    #no annoying stuff
    axis.title = element_blank(),
    legend.position = "none"
    )

new_plot
```

I'm pretty happy with this! I wish APTA colors were a little more dynamic, but this chart gets the main idea across faster.

## Export

Then, to export, a blank PNG is created and the plot is printed on top. This is different than `ggsave()` because of the custom fonts. I made the PNG the size of the Twitter cropped image so nothing would be lost.

```{r eval = FALSE}
# Export Plot

# {showtext} requires these steps, no ggsave available

png(filename = "new_plot.png",
    width = 600,
    height = 335)
new_plot
dev.off()

```



